### 2019-09-14 ic-journal/ccs.tex
text

\hl{Here we beliefly recall the core theory of CCS, up to the unique solution of
equation and contractions, with a focus on the syntactic differences with
other literature.}

@@

\ldots\}$ forming input and $\overline{\mbox{output}}$ labels (actions), plus a special invisible

@@

\hl{Notice the (minor) differences with other CCS literature regarding the above definition:}
\begin{enumerate}
  \item Some literature adopts possibly infinite sums over an index
    set ($\sum_{i\in I} P_i$) while we only consider binary (thus
    finite) sums, which is less expressive.
  \item Our restriction operator takes a set of (input) labels $L
    \subseteq \mathscr{L}$, while some CCS literature only restricts
    one label (for each restriction operator);
  \item Our relabeling operator takes a relabeling function

@@

    \{\tau\}$, which
    is capable to handle multiple actions including $\tau$. A valid
    relabeling function $\mathit{rf}$ must satisfy $\mathit{rf}(\tau) =

@@

    \mathscr{L}$); Same expressiveness
    with single-label relabeling.
  \item The recursion operator $\texttt{rec}$ has the same expressiveness with
    \emph{process constants} adopted in some literature. For instance, a
    CCS process $a.C$ with the constant $C \DSdefi b.C$ can be
    expressed as $a.(\texttt{rec}\, C. (b.C))$.
\end{enumerate}
Besides, we sometimes omit \hl{the} trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
\hl{A CCS process $P$ may \emph{derive} to another one, say $P'$, under
  an action $\mu$, denoted by $P \arr{\mu} P'$.}
The \hl{transitional semantics} of CCS processes is given by means of

@@

sums are in the form $a.P + b.Q$. \hl{(Otherwise we may call the sum
a \emph{direct} one, e.g. $P + Q$ if either $P$ or $Q$ are not in the
form of prefixes.)}

>>>>

\hlD{Here we recall the core theory of CCS.}

@@

\ldots\}$ \hlD{yielding} input and $\overline{\mbox{output}}$ labels (actions), plus a special invisible

@@

\hlD{In our presentation of CCS, 
the restriction operator takes a set of  labels $L
    \subseteq \mathscr{L}$, rather than a single one.  
The relabeling operator takes a relabeling function

@@

    \{\tau\}$, that can  handle multiple actions including $\tau$. A valid
    relabeling function $\mathit{rf}$ must however 
satisfy $\mathit{rf}(\tau) =

@@

    \mathscr{L}$).}
% \hl{Notice the (minor) differences with other CCS literature
% regarding the above definition:} 
% \begin{enumerate}
%   \item Some literature adopts possibly infinite sums over an index
%     set ($\sum_{i\in I} P_i$) while we only consider binary (thus
%     finite) sums, which is less expressive.
%   \item Our restriction operator takes a set of (input) labels $L
%     \subseteq \mathscr{L}$, while some CCS literature only restricts
%     one label (for each restriction operator);
%   \item Our relabeling operator takes a relabeling function
%     $\mathit{rf} \colon \mathscr{L} \cup \overline{\mathscr{L}} \cup
%     \{\tau\} \rightarrow \mathscr{L} \cup \overline{\mathscr{L}} \cup
%     \{\tau\}$, which
%     is capable to handle multiple actions including $\tau$. A valid
%     relabeling function $\mathit{rf}$ must satisfy $\mathit{rf}(\tau) =
%     \tau$ and $\forall l\in\mathscr{L} \cup
%     \overline{\mathscr{L}}.\, \mathit{rf}(\overline{l}) = 
%     \overline{\mathit{rf}(l)}$ (with $\bar{\bar l} = l$ for all $l \in
%     \mathscr{L}$); Same expressiveness
%     with single-label relabeling.
%   \item The recursion operator $\texttt{rec}$ has the same
%   expressiveness with  
%     \emph{process constants} adopted in some literature. For instance, a
%     CCS process $a.C$ with the constant $C \DSdefi b.C$ can be
%     expressed as $a.(\texttt{rec}\, C. (b.C))$.
% \end{enumerate}
We sometimes omit \hlD{a} trailing $\nil$, e.g., writing $a|b$ for $a.\nil |b .\nil$.
\hl{A CCS process $P$ may \hlD{evolve} to another one, say $P'$, under
  an action $\mu$, written by $P \arr{\mu} P'$.}
The \hlD{transition semantics} of CCS processes is given by means of

@@

sums are in the form $a.P + b.Q$. 
% \hl{(Otherwise we may call the sum
% a \emph{direct} one, e.g. $P + Q$ if either $P$ or $Q$ are not in the
% form of prefixes.)}

### 2019-09-14 ic-journal/equa.tex
Stage work on 2nd part
text

\hl{In the CCS syntax given in previous section, there is the syntactic
possibility that an agent variable $A$ occurs
outside the recursion operator of the same variable (or, equivalently, a
process constant $C$ is used but undefined). We say such a variable is
\emph{free}, while those variables wrapped by their recursions are
\emph{bound}.}
\hl{A CCS \emph{tem} (or \emph{expression}, i.e. those with valid syntax)
without any free variable is called a CCS \emph{process}.}
\hl{For instance, the variable $X$ is free in $a.X + \recu Y (b.Y)$, while $Y$
is bound. There is also the syntactic possibility that the same
variable $X$ is
both free and bound, e.g. in $a.X + \recu X (b.X)$.
This does not cause any ambiguity, but sometimes leads to more
complex proofs.} (c.f. Section~\ref{sec:multivariate} for more details.)

>>>>

\hl{In the CCS syntax, 
a recursion $\recu A  P$ acts as a binder for $A$ in the body $P$. 
This gives rise, in the expected manner, to the notions of 
\emph{free} and \emph{bound} recursion variables in a CCS expression.

 given in previous section, there is the syntactic
possibility that an agent variable $A$ occurs
outside the recursion operator of the same variable (or, equivalently, a
process constant $C$ is used but undefined). We say such a variable is
\emph{free}, while those variables wrapped by their recursions are
\emph{bound}.}
\hl{A CCS \emph{tem} (or \emph{expression}, i.e. those with valid syntax)
without any free variable is called a CCS \emph{process}.}
\hl{For instance, the variable $X$ is free in $a.X + \recu Y (b.Y)$, while $Y$
is bound. There is also the syntactic possibility that the same
variable $X$ is
both free and bound, e.g. in $a.X + \recu X (b.X)$.
This setting does not cause any ambiguity, but sometimes leads to more
complex proofs.} (c.f. Section~\ref{sec:multivariate} for more details.)

### 2019-09-14 ic-journal/expansion.tex
text

\hl{The \emph{strong} bisimulation and strong bisimilarity} ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with normal transitions $Q \arr \mu Q'$ (similar for the other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), \hl{i.e. from $P_1 \wb Q_1$ and $P_2 \wb Q_2$ one
  cannot obtain $P_1 + P_2 \wb Q_1 + Q_2$.}
For this, Milner introduced the concept of \emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, \hl{if} % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

\hl{The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma is usually not mentioned in CCS literature but plays an
important role in proving some key results in this paper:}
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
\hl{satisfy} the following properties \hl{(but not necessary $P\;\R\;Q$)}:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ \hl{implies that there is} $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
\hl{Then} $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

\hl{One basic property of rooted bisimiarity is that, it is (like $\sim$)
a congruence (thus also an equivalence relation)
w.r.t.~all CCS operators except for the Recursion and is the
coarsest congruence contained in bisimilarity:}
(c.f. Section~\ref{s:coarsest} for more details)
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

>>>>

\hlD{\emph{Strong} bisimulation and strong bisimilarity} ($\sim$)
can be obtained by replacing the weak transition $Q\Arcap\mu Q'$ in the above
definition with normal transitions $Q \arr \mu Q'$ (similar for the other case).
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums), \hlD{i.e.\  $P_1 \wb Q_1$ and $P_2 \wb Q_2$  does not imply
 $P_1 + P_2 \wb Q_1 + Q_2$.}
For this, Milner introduced the concept of \emph{observational congruence}, also called \emph{rooted
  bisimilarity}~\cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, \hl{if} % for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

\hlD{The above definition also brings up a proof technique for proving rooted
bisimilarity from a given bisimulation. The next
lemma plays an
important role in proving some key results in this paper:}
\begin{lemma}[Rooted bisimilarity by bisimulation]
\label{l:obsCongrByWeakBisim}
\hlD{Given a (weak) bisimulation $\RR$, suppose two processes $P$ and $Q$
satisfy the following properties}:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ \hl{implies that there is} $P'$ such that $P
   \Arr\mu P'$ and $P' \RR Q'$.
\end{enumerate}
\hl{Then} $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

\hlD{One basic property of rooted bisimiarity is congruence,
indeed
 the
coarsest congruence contained in bisimilarity
(c.f., Section~\ref{s:coarsest} for more details): }
\begin{theorem}
\label{t:rapproxCongruence}
$\rapprox$ is a congruence in CCS, and it is the
coarsest congruence contained in $\wb$.
\end{theorem}

### 2019-09-14 ic-journal/intro.tex
text

\hl{Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:}
the equations must be both \emph{(strongly) guarded} and \emph{sequential}. That is,
variables of the equations \hl{can only occur underneath visible prefixes and summations.}
One way \hl{to overcome} such limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency constraints on processes.
\hl{The unique solution of contractions is defined as with equations:}
any two solutions must be componentwise bisimilar.
The difference \hl{from} equations is \hl{at} the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem \hl{are
eliminated}. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of \emph{direct} sums (e.g.~$P + Q$) due to
the \hl{non-substitutivity of the contraction preorder in this case}.

The work \hl{behind} this paper is a
comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
\hl{Almost all results in Milner's CCS book}~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
\hl{bisimilarities} (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes \hl{Hennessy's Lemma} and \hl{Deng's Lemma},
 and two \hl{theorems} saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is \hl{also} an extensive experiment \hl{using} the HOL theorem prover with its
recent developments, including \hl{its coinductive relation-defining package}.

From the \hl{point of} view of CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, Sangiorgi's result~\cite{sangiorgi2017equations} has
placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
\hl{In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder} to
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
% rephrased: (TODO: still hard to read --Chun)
\hl{Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that uses the contraction preorder as an
auxiliary relation, and other unique solution results (e.g. the one
for rooted in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the `up-to' techniques).}
%
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

\hl{The univariate versions of unique-solution theorems use
$\lambda$-functions to present univariate CCS equations,
while the multivariate versions require a
careful and detailed formalisation of CCS equations with equation
variables and multivariate variable substitutions.
In contrast to most literature, we have adopted the same
type for both CCS equations and processes: those undefined constants
in CCS terms are treated as equation (free) variables, and a CCS
\emph{process} is a CCS term without equation variables. As a result,
most CCS results except the unique-solution theorems are valid for
both CCS processes and equations.} (c.f. Section~\ref{sec:multivariate}
for more deails.)

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Section~\ref{s:eq} discusses equations and contractions.
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
Section~\ref{s:for} highlights our formalisation in HOL4.
In Section~\ref{sec:multivariate}, \hl{we describe the formalisation of
multivariate versions of unique-solution theorems.}
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.

>>>>

\hl{Milner's unique-solution theorem for weak bisimilarity, however,
has severe syntactic limitations:}
the equations must be both \emph{(strongly) guarded} and \emph{sequential}. That is,
variables of the equations \hl{can only occur underneath visible prefixes and summations.}
One way \hl{to overcome} such limitations is to replace the equations
with special inequations called
\emph{contractions}~\cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency constraints on processes.
\hl{The unique solution of contractions is defined as with equations:}
any two solutions must be componentwise bisimilar.
The difference \hl{from} equations is \hlD{in} the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem \hl{are
eliminated}. One constraint that still remains
in~\cite{sangiorgi2017equations}
(where the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of %\emph{direct} 
\hlD{sums} (e.g.~$P + Q$) due to
the \hl{non-substitutivity of the contraction preorder in this case}.

 \hlD{This paper presents}  a
comprehensive formalisation of Milner's Calculus of Communicating
Systems (CCS) in the HOL theorem prover (HOL4),
with a focus towards the theory of unique solutions of equations and contractions.
\hlD{Many results in Milner's CCS book}~\cite{Mil89} are covered, since
the unique-solution theorems rely on a large number of fundamental results.
Indeed the formalisation encompasses all basic properties of strong and weak
\hl{bisimilarities} (e.g.~the fixed-point and substitutivity properties), the
basic properties of rooted bisimilarity and their algebraic laws.
Further extensions include several versions of ``bisimulation up to''
techniques, the expansion and contraction preorder.
Concerning rooted bisimilarity, the formalisation
includes \hl{Hennessy's Lemma} and \hl{Deng's Lemma},
 and two \hl{theorems} saying that rooted bisimilarity is the coarsest (largest)
 congruence contained in weak bisimilarity ($\wb$): one is classical
 with the hypothesis that no process uses up all labels;
the other without such hypothesis, essentially formalised van Glabbeek's proof~\cite{van2005characterisation}.
Similar results are also proved for the rooted contraction preorder.
In this respect, the work is \hl{also} an extensive experiment \hl{using} the HOL theorem prover with its
recent developments, including \hl{its coinductive relation-defining package}.

From the \hl{point of} view of the CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions.
In particular, \hlD{the existing result~\cite{sangiorgi2017equations} has
 limitations} on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
\hl{In this paper, the contraction-based proof technique is further
refined by moving from the contraction preorder} to
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
\hlD{constraints on the use the sum operator}.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures.
% rephrased: (TODO: still hard to read --Chun)
\hl{Such structural simililarities can be found, for instance, in the
following cases: the proofs that rooted bisimilarity and rooted
contraction are, respectively, the coarsest congruence contained in
weak bisimilarity and the coarsest precongruence contained in the
contraction preorder; the proofs about unique solution(s) of equations
for weak bisimilarity that uses the contraction preorder as an
auxiliary relation, and other unique solution results (e.g. the one
for rooted in which the auxiliary relation is rooted contraction); the
proofs about various forms of enhancements of the bisimulation proof
method (the `up-to' techniques).}
%
In these cases, when moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.

\finish{below: ``In contrast to most literature,...'' we need to
  support this with references}

\hlD{We first consider the case of a single equation (or contraction),
that we call the univariate case, then consider the multivariate
  case, where  more than one equation
  (or contraction) is possible.
The univariate  theorems use
$\lambda$-functions to present  CCS equations,
while the multivariate versions require a
more careful and delicate treatment of 
expressions with several 
variables and substitutions acting on them.
In contrast to most literature, we have adopted the same
type for both CCS equations and processes: those undefined constants
in CCS terms are treated as  (free) equation variables, and a CCS
\emph{process} is a CCS term without equation variables. 
This allows us a smoother move from the univariate case to the multivariate
one. See Section~\ref{sec:multivariate}
for  deails.}

@@

Section~\ref{ss:ccs} presents basic background materials on CCS,
including its syntax, operational semantics, bisimilarity and rooted
bisimilarity.
Section~\ref{s:eq} discusses equations and contractions.
Section~\ref{ss:new} presents rooted contraction and the related
unique-solution result for rooted bisimilarity.
Section~\ref{s:for} highlights our formalisation in HOL4 \hlDfor the
univariate case.}
\hlD{Section~\ref{sec:multivariate} describes the move to the
  multivariate case}. 
Finally,  Section~\ref{s:rel} and \ref{s:concl} discuss related work,
conclusions, and a few directions for future work.
